Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-15 iso-int/rc-iso-int.tex
minor

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
% properties and various constructions in it. However, for a
% computational model, there has been little focus on actually writing
% programs directly in {{langRevT}}. This paper aims to remedy that by
% presenting a systematic technique of expressing logically reversible
% small-step abstract machines in {{langRevT}}. In other words, this
% paper shows that once we have a logically reversible machine in a
% notation of our choice, expressing the machine as an isomorphic
% interpreter in {{langRevT}} can be done systematically and does not
% present any significant conceptual difficulties. This gives us a means
% of developing large {{langRevT}} programs with the ease of reasoning
% at the level of a conventional small-step semantics.  We develop
% several simple interpreters over numbers and addition, move onto tree
% traversal and finish with an interpreter for {{langRevT}} expressed in
% {{langRevT}}.
  In our previous work [RC'11, POPL'12], we developed a reversible
  programming language and established that every computation in it is a
  (partial) isomorphism that is reversible and that preserves
  information. The language is founded on type isomorphisms that have a clear
  categorical semantics but that are awkward as a notation for writing actual
  programs, especially recursive programs. This paper remedies this aspect by
  presenting a systematic technique for developing a large and expressive
  class of reversible recursive programs, that of logically reversible
  small-step abstract machines. In other words, this paper shows that once we
  have a logically reversible machine in a notation of our choice, expressing
  the machine as an isomorphic interpreter can be done systematically and
  does not present any significant conceptual difficulties. Concretely, we
  develop several simple interpreters over numbers and addition, move on to
  tree traversals, and finish with a meta-circular interpreter for our
  reversible language. Generally, this gives us a means of developing large
  reversible programs with the ease of reasoning at the level of a
  conventional small-step semantics.
\end{abstract}

@@

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and theSECD-H~\cite{lorenz});
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of tranforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
% properties and various constructions in it. However, for a
% computational model, there has been little focus on actually writing
% programs directly in {{langRevT}}. This paper aims to remedy that by
% presenting a systematic technique of expressing logically reversible
% small-step abstract machines in {{langRevT}}. In other words, this
% paper shows that once we have a logically reversible machine in a
% notation of our choice, expressing the machine as an isomorphic
% interpreter in {{langRevT}} can be done systematically and does not
% present any significant conceptual difficulties. This gives us a means
% of developing large {{langRevT}} programs with the ease of reasoning
% at the level of a conventional small-step semantics.  We develop
% several simple interpreters over numbers and addition, move onto tree
% traversal and finish with an interpreter for {{langRevT}} expressed in
% {{langRevT}}.
  In our previous work, we developed a reversible programming language and
  established that every computation in it is a (partial) isomorphism that is
  reversible and that preserves information. The language is founded on type
  isomorphisms that have a clear categorical semantics but that are awkward
  as a notation for writing actual programs, especially recursive
  programs. This paper remedies this aspect by presenting a systematic
  technique for developing a large and expressive class of reversible
  recursive programs, that of logically reversible small-step abstract
  machines. In other words, this paper shows that once we have a logically
  reversible machine in a notation of our choice, expressing the machine as
  an isomorphic interpreter can be done systematically and does not present
  any significant conceptual difficulties. Concretely, we develop several
  simple interpreters over numbers and addition, move on to tree traversals,
  and finish with a meta-circular interpreter for our reversible
  language. Generally, this gives us a means of developing large reversible
  programs with the ease of reasoning at the level of a conventional
  small-step semantics.
\end{abstract}

@@

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz});
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of tranforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}
